---
title: "NOAA Storm Data Analysis - Effects on Public Health and Commerce"
output: html_document
---

## Synopsis
The National Climatic Data Center(NCDC) regularly receives Storm Data from the National Weather Service(NWS).  The National Weather Service receives their information from a variety of sources, which include but are not limited to: county, state and federal emergency management officials, local law enforcement officials, skywarn spotters, NWS damage surveys, newspaper clipping services, the insurance industry and the general public.

An effort is made to use the best available information but due to time and resource constraints the information may be unverified by the NWS; therefore, the accuracy or validity of the information is not guaranteed.

* [Storm Data FAQ Page](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)  
* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  

## Data Proc

```{r download and unzip data}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, destfile = "repdata%2Fdata%2FStormData.csv.bz2", method = "curl")
stormData <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")
```

```{r Data Processing, message=FALSE, warning=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)

df <- summarize(stormData %>% group_by(EVTYPE),  FATALITIES = sum(FATALITIES))

top_10_health <- head(df[with(df, order(-FATALITIES)),], 10)

ggplot(data=top_10_health, aes(x = EVTYPE, y = FATALITIES)) + 
  geom_bar(position="dodge", stat="identity") + 
  coord_flip() + 
  scale_fill_grey() + 
  theme_bw()

```
#ggplot(df, aes(fatalities, group = EVTYPE, color = factor(EVTYPE))) + geom_line(size = 1)
              
              
### Get top health related costs
## Results



You can also embed plots, for example:

```{r, echo=FALSE}
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
