---
title: "NOAA Storm Data Analysis - Effects on Public Health and Commerce"
output: html_document
---

## Synopsis
The National Climatic Data Center(NCDC) regularly receives Storm Data from the National Weather Service(NWS).  The National Weather Service receives their information from a variety of sources, which include but are not limited to: county, state and federal emergency management officials, local law enforcement officials, skywarn spotters, NWS damage surveys, newspaper clipping services, the insurance industry and the general public.

An effort is made to use the best available information but due to time and resource constraints the information may be unverified by the NWS; therefore, the accuracy or validity of the information is not guaranteed.

This analysis will address the following questions:
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

* [Storm Data FAQ Page](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)  
* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  

## Data Processing

```{r}
#   Download data
#   Convert column names to lower case
#   Convert events (evtype) to lower case

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, destfile = "repdata%2Fdata%2FStormData.csv.bz2", method = "curl")
stormData <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")
names(stormData) <- tolower(names(stormData))
stormData$evtype <- tolower(stormData$evtype)
```

### Load libraries

```{r Data Processing, message=FALSE, warning=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(reshape2)
```

### What events are most harmful with respect to population health
#### Fatalities

```{r storm events causing fatalities, message=FALSE, warning=FALSE}
#   summarize stormData grouping by the event type and summing fatalities
#   plot the events that are responsible for the top 10 causes of death

dfFatalities <- summarize(stormData %>% group_by(evtype),  
                          fatalities = sum(fatalities))
top_10_fatalities <- head(dfFatalities[with(dfFatalities, 
                                            order(-fatalities)),], 10)

```

#### Injuries

```{r storm events causing injuries, message=FALSE, warning=FALSE}
#   summarize stormData grouping by the event type and summing injuries
#   plot the events that are responsible for the top 10 causes of injury

dfInjuries <- summarize(stormData %>% group_by(evtype),  injuries = sum(injuries))
top_10_injuries <- head(dfInjuries[with(dfInjuries, order(-injuries)),], 10)

```

### What types of events have the greatest economic consequences



```{r}
# function designed to convert exponential codes into numeric values
# codes not found in documentation added
exponent_translate <- function(exp) {
    # k -> thousand, m -> million, b -> billion
    if (exp %in% c('k', 'K'))
        return(3)
    else if (exp %in% c('m', 'M'))
        return(6)
    else if (exp %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(exp))) # if a digit
        return(as.numeric(exp))
    else if (exp %in% c('', '-', '?', '+'))
        return(0)
    #else {
    #    stop("Invalid exponent value.")
    #}
}
```

```{r storm events with greatest economic consequences}
#   filter stormData for propdmg, propdmgexp, cropdmg, cropdmgexp 
#   translate the propdmgexp and cropdmgexp, applied to property and crop damages 
#   summarizing by event 
#   divide costs by 1,000,000
#   add column totaldmg representing contributions from property and crop damages
#   summarize results grouping by the event type 
#   generate a top 10 event list based on totaldmg
#   reshape into tidy data set
#   plot the events that are responsible for the top 10 causes of injury

stormData_DMG <- select(stormData, evtype, propdmg, cropdmg, propdmgexp, cropdmgexp)

prop_dmg_exp <- sapply(stormData_DMG$propdmgexp, FUN=exponent_translate)
stormData_DMG$propdmg <- stormData_DMG$propdmg * (10 ** prop_dmg_exp)

crop_dmg_exp <- sapply(stormData_DMG$cropdmgexp, FUN=exponent_translate)
stormData_DMG$cropdmg <- stormData_DMG$cropdmg * (10 ** crop_dmg_exp)

stormData_DMG_Event <- summarize(stormData_DMG %>% group_by(evtype), 
                            propdmg=sum(propdmg), cropdmg=sum(cropdmg))
stormData_DMG_Event$propdmg <- stormData_DMG_Event$propdmg / 1000000
stormData_DMG_Event$cropdmg <- stormData_DMG_Event$cropdmg / 1000000

stormData_DMG_Event$totaldmg <- stormData_DMG_Event$propdmg + stormData_DMG_Event$cropdmg 

top_10_DMG_Event <- head(stormData_DMG_Event[with(stormData_DMG_Event, order(-totaldmg)),], 10)

tidy_top_10 <- melt(top_10_DMG_Event, id.vars = "evtype", variable.name = "damage")

```

## Results
```
1. Across the United States, which types of events are most harmful with respect to population health?
    -Top 10 storm events resulting in fatalities
```
```{r}
ggplot(data=top_10_fatalities, aes(x = evtype, y = fatalities)) + 
    ggtitle("Top 10 Storm Related Fatalities") +
    geom_bar(position="dodge", stat="identity") + 
    coord_flip() + 
    scale_fill_grey() + 
    theme_bw() +
    theme(axis.title.y=element_blank())
```
```
1. Across the United States, which types of events are most harmful with respect to population health?
    -Top 10 storm events resulting in injuries
```
```{r}
ggplot(data=top_10_injuries, aes(x = evtype, y = injuries)) + 
    ggtitle("Top 10 Storm Related Injuries") +
    geom_bar(position="dodge", stat="identity") + 
    coord_flip() + 
    scale_fill_grey() + 
    theme_bw() +
    theme(axis.title.y=element_blank()) 
```
```
2. Across the United States, which types of events have the greatest economic consequences? 
```
```{r}
ggplot(data=tidy_top_10, aes(x = evtype, y = value, fill=damage)) + 
    ggtitle("Top 10 Storm Related Economic Costs\n Period 1950 - Nov 2011") +
    geom_bar(position="dodge", stat="identity") + 
    coord_flip() + 
    theme(axis.title.y=element_blank()) +
    ylab("Costs in $1,000,000") +
    scale_fill_manual(values=c("grey70", "grey85", "black"),
                        name = "damage",
                        breaks = c("propdmg", "cropdmg", "totaldmg"),
                        labels = c("property", "crop", "total")) 

```